---
title: Animation
subtitle: Lecture 25
format: revealjs
---

# Warm up

## Announcements {.smaller}

## Setup {.smaller}

```{r}
#| label: setup
#| message: false
# load packages
library(tidyverse)
library(gganimate)
library(gifski)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7, # 7" width
  fig.asp = 0.618, # the golden ratio
  fig.retina = 3, # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300 # higher dpi, sharper image
)
```

# Cumulative inflation data

## Data source {.smaller}

Consumer Price Index (CPI) data from [FRED](https://fred.stlouisfed.org/) (Federal Reserve Economic Data):

| Country | Series ID | Description |
|---------|-----------|-------------|
| USA | [CPIAUCSL](https://fred.stlouisfed.org/series/CPIAUCSL) | CPI for All Urban Consumers: All Items |
| Turkiye | [TURCPIALLMINMEI](https://fred.stlouisfed.org/series/TURCPIALLMINMEI) | CPI: All Items for Turkey |
| Australia | [AUSCPIALLQINMEI](https://fred.stlouisfed.org/series/AUSCPIALLQINMEI) | CPI: All Items for Australia |
| Europe | [CP0000EZ19M086NEST](https://fred.stlouisfed.org/series/CP0000EZ19M086NEST) | HICP: All Items for Euro Area |
| Japan | [FPCPITOTLZGJPN](https://fred.stlouisfed.org/series/FPCPITOTLZGJPN) | Inflation, consumer prices (annual %) |

All series use **annual averages** with data from 1996-2024.

## Methodology {.smaller}

**Goal:** Calculate cumulative percentage change in prices since 1996, rebased to 0% at inception.

**For countries with CPI index levels** (USA, Turkiye, Australia, Europe):

$$\text{Cumulative Inflation}_t = \left( \frac{\text{CPI}_t}{\text{CPI}_{1996}} - 1 \right) \times 100$$

**For Japan** (only annual inflation rates available):

1. Construct index starting at 100 in 1996
2. Compound forward: $\text{Index}_t = \text{Index}_{t-1} \times (1 + r_t / 100)$
3. Calculate cumulative change: $\text{Cumulative}_t = \text{Index}_t - 100$

## Construct the data - download {.smaller}

```{r}
#| label: download-cpi-data
#| eval: false
library(tidyverse)

# FRED API URLs for annual average CPI data (1996-2024)
base_url <- "https://fred.stlouisfed.org/graph/fredgraph.csv"

# Download CPI index levels
usa <- read_csv(glue::glue("{base_url}?id=CPIAUCSL&cosd=1996-01-01&coed=2024-12-31&fq=Annual&fam=avg"))
turkiye <- read_csv(glue::glue("{base_url}?id=TURCPIALLMINMEI&cosd=1996-01-01&coed=2024-12-31&fq=Annual&fam=avg"))
australia <- read_csv(glue::glue("{base_url}?id=AUSCPIALLQINMEI&cosd=1996-01-01&coed=2024-12-31&fq=Annual&fam=avg"))
europe <- read_csv(glue::glue("{base_url}?id=CP0000EZ19M086NEST&cosd=1996-01-01&coed=2024-12-31&fq=Annual&fam=avg"))

# Japan: only annual inflation rates available
japan_rates <- read_csv(glue::glue("{base_url}?id=FPCPITOTLZGJPN&cosd=1996-01-01&coed=2024-12-31&fq=Annual&fam=avg"))
```

## Construct the data - calculate {.smaller}

```{r}
#| label: calculate-cumulative
#| eval: false
# Function to calculate cumulative inflation from index levels
calc_cumulative <- function(df, value_col) {
  df |>
    mutate(
      year = year(observation_date),
      cumulative = ({{ value_col }} / first({{ value_col }}) - 1) * 100
    ) |>
    select(year, cumulative)
}

# Calculate for countries with index levels
usa_cum <- calc_cumulative(usa, CPIAUCSL) |> rename(usa = cumulative)
turkiye_cum <- calc_cumulative(turkiye, TURCPIALLMINMEI) |> rename(turkiye = cumulative)
australia_cum <- calc_cumulative(australia, AUSCPIALLQINMEI) |> rename(australia = cumulative)
europe_cum <- calc_cumulative(europe, CP0000EZ19M086NEST) |> rename(europe = cumulative)

# Japan: construct index from annual inflation rates, then calculate cumulative
japan_cum <- japan_rates |>
  mutate(year = year(observation_date)) |>
  arrange(year) |>
  mutate(
    index = accumulate(FPCPITOTLZGJPN[-1], ~ .x * (1 + .y / 100), .init = 100),
    japan = index - 100
  ) |>
  select(year, japan)
```

## Construct the data - combine {.smaller}

```{r}
#| label: combine-data
#| eval: false
# Combine all countries
inflation <- usa_cum |>
  left_join(turkiye_cum, by = "year") |>
  left_join(australia_cum, by = "year") |>
  left_join(europe_cum, by = "year") |>
  left_join(japan_cum, by = "year") |>
  mutate(across(where(is.numeric) & !year, ~ round(.x, 2)))

# Save to CSV
write_csv(inflation, "data/cpi_cumulative_all.csv")
```

## Load the data {.smaller}

```{r}
#| label: load-inflation-data
#| message: false
inflation <- read_csv("data/cpi_cumulative_all.csv")
inflation
```

## Results summary {.smaller}

```{r}
#| label: inflation-summary
inflation |>
  filter(year == 2024) |>
  pivot_longer(cols = -year, names_to = "country", values_to = "cumulative_pct") |>
  arrange(desc(cumulative_pct))
```

Key findings:

- **Turkiye:** 56,149% - extreme hyperinflation, prices 562x higher than 1996
- **Australia & USA:** ~100-108% - prices roughly doubled
- **Europe:** ~77% - more moderate inflation
- **Japan:** ~13% - remarkably flat, experienced deflation in many years

# Animated visualization

## Reshape data for plotting {.smaller}

```{r}
#| label: reshape-data
inflation_long <- inflation |>
  pivot_longer(
    cols = -year,
    names_to = "country",
    values_to = "cumulative_pct"
  ) |>
  mutate(
    country = case_when(
      country == "usa" ~ "USA",
      country == "turkiye" ~ "Turkiye",
      country == "australia" ~ "Australia",
      country == "europe" ~ "Europe",
      country == "japan" ~ "Japan"
    ),
    country = fct_reorder2(country, year, cumulative_pct)
  )

inflation_long
```

## Static plot (excluding Turkiye) {.smaller}

```{r}
#| label: static-plot
inflation_long |>
  filter(country != "Turkiye") |>
  ggplot(aes(x = year, y = cumulative_pct, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Cumulative inflation since 1996",
    x = "Year",
    y = "Cumulative inflation (%)",
    color = "Country"
  )
```

## Animated plot (excluding Turkiye) {.smaller}

```{r}
#| label: animated-plot
#| warning: false
inflation_long |>
  filter(country != "Turkiye") |>
  ggplot(aes(x = year, y = cumulative_pct, color = country, group = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Cumulative inflation since 1996",
    subtitle = "Year: {frame_along}",
    x = "Year",
    y = "Cumulative inflation (%)",
    color = "Country"
  ) +
  transition_reveal(year)
```

## Animated plot with Turkiye (log scale) {.smaller}

```{r}
#| label: animated-plot-all
#| warning: false
inflation_long |>
  ggplot(aes(x = year, y = cumulative_pct + 1, color = country, group = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10(labels = scales::label_comma(suffix = "%")) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Cumulative inflation since 1996 (log scale)",
    subtitle = "Year: {frame_along}",
    x = "Year",
    y = "Cumulative inflation (%, log scale)",
    color = "Country"
  ) +
  transition_reveal(year)
```

## Racing bar chart {.smaller}

```{r}
#| label: bar-race
#| warning: false
inflation_long |>
  filter(country != "Turkiye") |>
  ggplot(aes(x = cumulative_pct, y = country, fill = country)) +
  geom_col() +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Cumulative inflation since 1996",
    subtitle = "Year: {closest_state}",
    x = "Cumulative inflation (%)",
    y = NULL
  ) +
  transition_states(year, transition_length = 2, state_length = 1) +
  ease_aes("cubic-in-out")
```

## Turkiye: Expanding y-axis animation {.smaller}

```{r}
#| label: turkiye-expanding
#| warning: false
inflation_long |>
  filter(country == "Turkiye") |>
  ggplot(aes(x = year, y = cumulative_pct)) +
  geom_line(linewidth = 1.2, color = "#E41A1C") +
  geom_point(size = 3, color = "#E41A1C") +
  geom_text(
    aes(label = scales::comma(round(cumulative_pct), suffix = "%")),
    hjust = -0.2, vjust = 0.5, size = 4
  ) +
  scale_y_continuous(labels = scales::label_comma(suffix = "%")) +
  labs(
    title = "Turkiye: Cumulative inflation since 1996",
    subtitle = "Year: {frame_along}",
    x = "Year",
    y = "Cumulative inflation (%)"
  ) +
  transition_reveal(year) +
  view_follow(fixed_x = TRUE)
```

## All countries with expanding y-axis {.smaller}

```{r}
#| label: all-expanding
#| warning: false
inflation_long |>
  ggplot(aes(x = year, y = cumulative_pct, color = country, group = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::label_comma(suffix = "%")) +
  labs(
    title = "Cumulative inflation since 1996",
    subtitle = "Year: {frame_along}",
    x = "Year",
    y = "Cumulative inflation (%)",
    color = "Country"
  ) +
  transition_reveal(year) +
  view_follow(fixed_x = TRUE)
```

## All countries: Expanding axes on dark background {.smaller}

```{r}
#| label: all-expanding-dark
#| warning: false
inflation_long |>
  ggplot(aes(x = year, y = cumulative_pct, color = country, group = country)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c(
      "USA" = "#00BFFF",
      "Turkiye" = "#FF4500",
      "Australia" = "#00FF7F",
      "Europe" = "#FFD700",
      "Japan" = "#FF69B4"
    )
  ) +
  scale_y_continuous(labels = scales::label_comma(suffix = "%")) +
  labs(
    title = "Cumulative inflation since 1996",
    subtitle = "Year: {frame_along}",
    x = "Year",
    y = "Cumulative inflation (%)",
    color = "Country"
  ) +
  theme(
    plot.background = element_rect(fill = "black", color = "black"),
    panel.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    text = element_text(color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", size = 16, face = "bold"),
    plot.subtitle = element_text(color = "gray80", size = 12),
    legend.background = element_rect(fill = "black"),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  ) +
  transition_reveal(year) +
  view_follow(fixed_x = FALSE, fixed_y = FALSE)
```

