---
title: 2020 Durham City and County Resident Survey
subtitle: AE 06
execute: 
  eval: false
---

# Setup

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(scales)

durham <- read_csv("data/durham-2020.csv")
```

## Data prep

First, clean/tidy/reshape the data so that it can be used to create the improved plots from the slides.

Some notes:

-   Drop any responses with `NA`s

-   The questions of interest are the only ones with `_3_` in their names:

    -   `overall_quality_of_services_3_01`: Overall quality of services provided by the city

    -   `overall_quality_of_services_3_02`: Overall quality of services provided by the county

    -   `overall_appearance_of_durham_3_03`: Overall appearance of Durham

    -   `overall_management_of_develo_3_04`: Overall management of development and growth

    -   `overall_image_of_durham_3_05`: Overall image of Durham

    -   `overall_quality_of_life_in_d_3_06`: Overall quality of life in Durham

    -   `overall_quality_of_life_in_y_3_07`: Overall quality of life in your neighborhood

    -   `overall_value_you_receive_fo_3_08`: Overall value you receive for your local property taxes

-   Responses:

    -   1: Very Dissatisfied

    -   2: Dissatisfied

    -   3: Neutral

    -   4: Satisfied

    -   5: Very Satisfied

Save the cleaned/tidied/reshaped data as `durham_overall_quality`.

```{r}
#| label: likert-data-prep
# add code here
```

## Starting point

Then, make the simple stacked bar chart in Step 1.

```{r}
#| label: likert-100-step-1
#| fig-asp: 0.5
#| fig-width: 10
# add code here
```

## 100% bar chart

Improve the visualization above to the 100% bar chart in Step 2.

```{r}
#| label: likert-100-step-2
#| fig-asp: 0.5
#| fig-width: 10
# add code here
```

## Diverging bar chart

Improve the visualization further to the diverging bar chart in Step 3.

```{r}
#| label: likert-100-step-3
#| fig-asp: 0.5
#| fig-width: 10
# add code here
```
