---
title: Data visualization with Python |
subtitle: Lecture 22
format: revealjs
filters: 
  - ../remove-fmt-skip.lua
---

# Warm up

## Announcements {.smaller}

- ...

## From last time: Generative art resources

-   R packages:

    -   [aRtsy](https://koenderks.github.io/aRtsy/) by Koen Derks
    -   [generativeart](https://github.com/cutterkom/generativeart) by Katharina Brunner
    -   [flametree](https://flametree.djnavarro.net/) and [jasmines](https://jasmines.djnavarro.net/) bu Danielle Navarro
    -   [ambient](https://ambient.data-imaginist.com/) and [particles](https://github.com/thomasp85/particles) by Thomas Lin Pedersen

-   More aRtists: [rtistry art gallery](https://ijeamaka-a.shinyapps.io/rtistry_gallery/) by Ijeamaka Anyene

-   A whole course on Generative Art by Danielle Navarro: [https://art-from-code.netlify.app](https://art-from-code.netlify.app/)

## Setup {.smaller}

- R:

```{r}
#| message: false
library(tidyverse)
```

- Python:

```{python}
import polars as pl
from plotnine import ggplot, aes, geom_point, geom_segment, labs, theme, theme_minimal
```

# Data visualization with Python

## Spot the difference

::: {.columns}

::: {.column}

**Plot A:**

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 4
#| echo: false
df_r <- read_csv("data/mean-area-decade.csv")

ggplot(
  df_r,
  aes(x = mean_area, y = decade_built_cat)
) +
  geom_segment(
    aes(
      x = 0,
      xend = mean_area,
      y = decade_built_cat,
      yend = decade_built_cat
    )
  ) +
  geom_point(size = 5, color = "lightblue") +
  labs(
    x = "Mean area (square feet)",
    y = "Decade built",
    title = "Mean area of houses in Duke Forest, by decade built"
  ) +
  theme_minimal()
```

:::

::: {.column}

**Plot B:**

```{python}
#| echo: false
df_py = pl.read_csv("data/mean-area-decade.csv")

(
    ggplot(df_py, aes(x="mean_area", y="decade_built_cat"))
    + geom_segment(
        aes(x=0, xend="mean_area", y="decade_built_cat", yend="decade_built_cat")
    )
    + geom_point(size=5, color="lightblue")
    + labs(
        x="Mean area (square feet)",
        y="Decade built",
        title="Mean area of houses in Duke Forest, by decade built",
    )
    + theme_minimal()
    + theme(figure_size=(8, 4))
)
```

:::

:::

## Spot the difference

::: {.columns}

::: {.column}

**Plot A:**

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 4
#| echo: false
df_r <- read_csv("data/mean-area-decade.csv")

ggplot(
  df_r,
  aes(x = mean_area, y = decade_built_cat)
) +
  geom_segment(
    aes(
      x = 0,
      xend = mean_area,
      y = decade_built_cat,
      yend = decade_built_cat
    )
  ) +
  geom_point(size = 5, color = "lightblue") +
  labs(
    x = "Mean area (square feet)",
    y = "Decade built",
    title = "Mean area of houses in Duke Forest, by decade built"
  ) +
  theme_minimal()
```

::: center-align

{{< fa brands r-project size=Huge >}}

:::

:::

::: {.column}

**Plot B:**

```{python}
#| echo: false
df_py = pl.read_csv("data/mean-area-decade.csv")

(
    ggplot(df_py, aes(x="mean_area", y="decade_built_cat"))
    + geom_segment(
        aes(x=0, xend="mean_area", y="decade_built_cat", yend="decade_built_cat")
    )
    + geom_point(size=5, color="lightblue")
    + labs(
        x="Mean area (square feet)",
        y="Decade built",
        title="Mean area of houses in Duke Forest, by decade built",
    )
    + theme_minimal()
    + theme(figure_size=(8, 4))
)
```

::: center-align

{{< fa brands python size=Huge >}}

:::

:::

:::



## Overview

::: incremental

- Goal: 
  - Get a taste of data visualization with Python with `plotnine` -- a Python data visualization package based on the grammar of graphics and inspired by `ggplot2`.
  - Not a comprehensive introduction to Python or even to data visualization with Python.

- Approach: 
  - We will cover the basics of `plotnine` and how to create different types of plots.
  - We will use `polars` to read in data from CSV files, but we will **not** cover how to prepare your data for visualization in Python.

:::

## But...

<br> <br>

::: large

[there are some Python details we can't avoid...]{.hand}

:::

# Package management <br> in Python with `uv`

## Python packaging landscape

Python's packaging ecosystem has historically been fragmented:

* Multiple tools: `pip`, `virtualenv`, `venv`, `conda`, `poetry`, `pipenv`, etc.
* Multiple config files: `requirements.txt`, `setup.py`, `pyproject.toml`, etc.
* Version management often handled separately (`pyenv`)

. . .

**uv** is a modern tool that aims to unify these concerns with a fast, Rust-based implementation.


## What is uv?

**uv** is a Python package and project manager developed by [Astral](https://astral.sh).

<br/>

Key features:

* Extremely fast (10-100x faster than pip)
* Manages Python versions
* Creates and manages virtual environments
* Installs packages
* Handles project dependencies via `pyproject.toml`
* Drop-in replacement for `pip` and `virtualenv`
* Directly supported by Positron and Reticulate

## Installing uv

**uv** is already installed on the departmental servers, for local installs:


On MacOS/Linux:

::: {.small}

```bash
curl -LsSf https://astral.sh/uv/install.sh | sh
```

:::

or with homebrew:

::: {.small}

```bash
brew install uv
```

:::

or with pip / pipx

::: {.small}

```bash
pipx install uv
pip install uv
```

:::


## Verify installation

Once installed you should be able to run the following,

```{bash}
uv --version
```

. . .

<br/>

As long as you have something higher than `0.9.*` you should be fine.

## Python versions

```{bash}
uv python list
```

## Managing Python versions

uv can install and manage multiple Python versions,

```{bash}
uv python install 3.14
```

```{bash}
uv python pin 3.14
```

. . .

The pinned version is stored in `~/.python-version` and will be used automatically.


## Initializing a project

Use `uv init` to create a new project,

::: {.small}

```bash
mkdir my-project
cd my-project
uv init
```

```
Initialized project `my-project`
```

```bash
ls -la
```

```
total 32
drwxr-xr-x   8 mine  staff   256 Apr  6 00:22 .
drwx------@ 55 mine  staff  1760 Apr  6 00:22 ..
drwxr-xr-x@  9 mine  staff   288 Apr  6 00:22 .git
-rw-r--r--@  1 mine  staff   109 Apr  6 00:22 .gitignore
-rw-r--r--@  1 mine  staff     5 Apr  6 00:22 .python-version
-rw-r--r--@  1 mine  staff    88 Apr  6 00:22 main.py
-rw-r--r--@  1 mine  staff   156 Apr  6 00:22 pyproject.toml
-rw-r--r--@  1 mine  staff     0 Apr  6 00:22 README.md
```

:::

. . .

This creates a `pyproject.toml`, a sample `main.py` script, and basic git infrastructure. Generally, we only really care about the `pyproject.toml` which we can exclusively generate via `uv init --bare`.

## `pyproject.toml`

Modern project metadata file, tracks python version and package dependencies among other details.

```toml
[project]
name = "my-project"
version = "0.1.0"
description = "Add your description here"
readme = "README.md"
requires-python = ">=3.14"
dependencies = []
```

## Adding dependencies

Once we have our project setup we can add (and install) dependencies directly via uv. `uv add` updates `pyproject.toml` and installs the package (creating a venv if needed).

::: {.columns .xsmall}

::: {.column}

```bash
uv add plotnine
```

```
Using CPython 3.14.3
Creating virtual environment at: .venv
Resolved 19 packages in 188ms
Installed 17 packages in 134ms
 + contourpy==1.3.3
 + cycler==0.12.1
 + fonttools==4.62.1
 + kiwisolver==1.5.0
 + matplotlib==3.10.8
 + mizani==0.14.4
 + numpy==2.4.4
 + packaging==26.0
 + pandas==3.0.2
 + patsy==1.0.2
 + pillow==12.2.0
 + plotnine==0.15.3
 + pyparsing==3.3.2
 + python-dateutil==2.9.0.post0
 + scipy==1.17.1
 + six==1.17.0
 + statsmodels==0.14.6
```

:::

::: {.column .fragment}

```bash
uv add polars
```

```
Resolved 21 packages in 142ms
Installed 2 packages in 6ms
 + polars==1.39.3
 + polars-runtime-32==1.39.3
```

:::

::: 

## Updated `pyproject.toml`

```toml
[project]
name = "my-project"
version = "0.1.0"
description = "Add your description here"
readme = "README.md"
requires-python = ">=3.14"
dependencies = [
    "plotnine>=0.15.3",
    "polars>=1.39.3",
]
```

## Virtual environments

Virtual environments isolate project dependencies from the system Python and other projects. Packages are installed in a local folder in your project.

As we just saw, using `uv add` will create a new virtual environment in `.venv` by default if there is not an existing venv.

## Activating environments in Positron with uv

Positron automatically detects virtual environments in your project directory. When you open a folder containing a `.venv` directory (created by uv), Positron will:

* Detect the environment and offer to use it
* Show the active Python interpreter in the status bar
* Use the environment for the Python console and when running scripts

If not automatically detected, you can manually select the interpreter via the Command Palette (`Cmd+Shift+P` / `Ctrl+Shift+P`) and searching for "Python: Select Interpreter".

## `uv sync`

Since the `.venv` folder is system specific (and large) it is not typically committed to git. Instead you will likely clone a repository that just has a `pyproject.toml` file.

Use `uv sync` to construct the venv and install all dependencies for the project 

## Listing packages

::: {.small}

```bash
uv pip list
```

```
Package           Version
----------------- -----------
contourpy         1.3.3
cycler            0.12.1
fonttools         4.62.1
kiwisolver        1.5.0
matplotlib        3.10.8
mizani            0.14.4
numpy             2.4.4
packaging         26.0
pandas            3.0.2
patsy             1.0.2
pillow            12.2.0
plotnine          0.15.3
polars            1.39.3
polars-runtime-32 1.39.3
pyparsing         3.3.2
python-dateutil   2.9.0.post0
scipy             1.17.1
six               1.17.0
statsmodels       0.14.6
```

```bash
uv pip show plotnine
```

```
Name: plotnine
Version: 0.15.3
Location: /Users/mine/Desktop/my-project/.venv/lib/python3.14/site-packages
Requires: matplotlib, mizani, numpy, pandas, scipy, statsmodels
Required-by:
```

:::

`uv pip install` and `uv pip freeze` are also available as lower-level commands for working outside of a project context or with `requirements.txt` files.


## Dependency tree

::: {.xsmall}

```bash
uv tree
```

```
Resolved 21 packages in 18ms
my-project v0.1.0
в”њв”Ђв”Ђ plotnine v0.15.3
в”‚   в”њв”Ђв”Ђ matplotlib v3.10.8
в”‚   в”‚   в”њв”Ђв”Ђ contourpy v1.3.3
в”‚   в”‚   в”‚   в””в”Ђв”Ђ numpy v2.4.4
в”‚   в”‚   в”њв”Ђв”Ђ cycler v0.12.1
в”‚   в”‚   в”њв”Ђв”Ђ fonttools v4.62.1
в”‚   в”‚   в”њв”Ђв”Ђ kiwisolver v1.5.0
в”‚   в”‚   в”њв”Ђв”Ђ numpy v2.4.4
в”‚   в”‚   в”њв”Ђв”Ђ packaging v26.0
в”‚   в”‚   в”њв”Ђв”Ђ pillow v12.2.0
в”‚   в”‚   в”њв”Ђв”Ђ pyparsing v3.3.2
в”‚   в”‚   в””в”Ђв”Ђ python-dateutil v2.9.0.post0
в”‚   в”‚       в””в”Ђв”Ђ six v1.17.0
в”‚   в”њв”Ђв”Ђ mizani v0.14.4
в”‚   в”‚   в”њв”Ђв”Ђ numpy v2.4.4
в”‚   в”‚   в”њв”Ђв”Ђ pandas v3.0.2
в”‚   в”‚   в”‚   в”њв”Ђв”Ђ numpy v2.4.4
в”‚   в”‚   в”‚   в””в”Ђв”Ђ python-dateutil v2.9.0.post0 (*)
в”‚   в”‚   в””в”Ђв”Ђ scipy v1.17.1
в”‚   в”‚       в””в”Ђв”Ђ numpy v2.4.4
в”‚   в”њв”Ђв”Ђ numpy v2.4.4
в”‚   в”њв”Ђв”Ђ pandas v3.0.2 (*)
в”‚   в”њв”Ђв”Ђ scipy v1.17.1 (*)
в”‚   в””в”Ђв”Ђ statsmodels v0.14.6
в”‚       в”њв”Ђв”Ђ numpy v2.4.4
в”‚       в”њв”Ђв”Ђ packaging v26.0
в”‚       в”њв”Ђв”Ђ pandas v3.0.2 (*)
в”‚       в”њв”Ђв”Ђ patsy v1.0.2
в”‚       в”‚   в””в”Ђв”Ђ numpy v2.4.4
в”‚       в””в”Ђв”Ђ scipy v1.17.1 (*)
в””в”Ђв”Ђ polars v1.39.3
    в””в”Ђв”Ђ polars-runtime-32 v1.39.3
(*) Package tree already displayed
```

:::


## Common workflows

New project setup:

::: {.small}

```bash
mkdir my-project # or use Positron's new project wizard
cd my-project    # and navigate to the project folder
uv init --bare
uv add plotine polars
```

:::

. . .

<br/>

Clone existing project:

::: {.small}

```bash
git clone <repo-url> # or use Positron's git integration
cd <repo>            # and navigate to the project folder
uv sync
```

:::

## Let's give it a try!

::: task
Go to `ae-16` and let's make a simple plot with `plotnine`!
:::

# Introduction to plotnine

## What is plotnine?

**plotnine** is a Python visualization library that implements the **grammar of graphics**.

::: incremental

- Based on the same principles as ggplot2 in R
- Enables data visualization through composable, layered components
- Maps data to visual properties systematically
- Works with Pandas and Polars DataFrames

:::

## The grammar of graphics

A plot is built from **layers** of components:

| Component | Description |
|-----------|-------------|
| **Data** | The dataset to visualize |
| **Aesthetics** | Mappings from data to visual properties |
| **Geoms** | Geometric objects that represent data |
| **Scales** | Control how data values map to visual values |
| **Facets** | Split data into multiple subplots |
| **Coords** | Coordinate system for the plot |
| **Themes** | Control non-data visual elements |

## Data

All plots begin with passing data to `ggplot()`:

```{python}
#| eval: false
from plotnine import ggplot

(
    ggplot(data=my_dataframe)
)
```

. . .

::: callout-tip

Plotnine works best with **tidy data**:

- Each variable is a column
- Each observation is a row
- Each type of observational unit is a table

:::

## Aesthetic mappings {.smaller}

The `aes()` function maps data columns to visual properties:

```{python}
#| eval: false
from plotnine import ggplot, aes

ggplot(df, aes(x="column1", y="column2"))
```

. . .

<br/>

Common aesthetic mappings:

| Aesthetic | Description |
|-----------|-------------|
| `x`, `y` | Position on axes |
| `color` | Color of points/lines |
| `fill` | Fill color of shapes |
| `size` | Size of points |
| `shape` | Shape of points |
| `alpha` | Transparency |

## Geometric objects (geoms)

```{python}
#| eval: false
from plotnine import ggplot, aes, geom_point

(
    ggplot(df, aes(x="x_var", y="y_var")) 
    + geom_point()
)
```

. . .

<br/>

::: {.callout-note}
In Python, wrap the entire plot expression in parentheses `()` to allow line breaks.
:::

## Geometric objects (geoms) {.smaller}

Geoms determine how data is visually represented:

| Geom | Description |
|------|-------------|
| `geom_point()` | Scatter plot |
| `geom_line()` | Line plot |
| `geom_bar()` | Bar chart |
| `geom_histogram()` | Histogram |
| `geom_boxplot()` | Box plot |
| `geom_smooth()` | Smoothed line |
| `geom_text()` | Text labels |
| `geom_segment()` | Line segments |
| `geom_area()` | Area plot |
| `geom_density()` | Density plot |

## Layering with `+`

Components are combined using the `+` operator:

```{python}
#| eval: false
from plotnine import ggplot, aes, geom_point, geom_smooth

(
    ggplot(df, aes(x="x_var", y="y_var"))
    + geom_point()
    + geom_smooth(method="lm")
)
```

. . .

<br/>

::: {.callout-note}
In Python, move the `+` to the start of the line and add a line break before `+`.
:::

## Scales {.smaller}

Scales customize how data values map to visual values.

Naming pattern: `scale_<aesthetic>_<type>`

. . .

```{python}
#| eval: false
from plotnine import ggplot, aes, geom_point, scale_color_continuous

(
    ggplot(df, aes(x="x_var", y="y_var", color="z_var"))
    + geom_point()
    + scale_color_continuous(cmap_name="viridis")
)
```

. . .

Common scale functions:

- `scale_x_continuous()`, `scale_y_continuous()` - continuous axes
- `scale_x_log10()`, `scale_y_log10()` - log-transformed axes
- `scale_color_manual()`, `scale_fill_manual()` - custom colors
- `scale_color_brewer()`, `scale_fill_brewer()` - ColorBrewer palettes

## Facets {.smaller}

Facets split data into multiple subplots:

::: {.columns}

::: {.column width="50%"}

**`facet_wrap()`** - wraps panels into rows

```{python}
#| eval: false
from plotnine import facet_wrap

(
    ggplot(df, aes(x="x", y="y"))
    + geom_point()
    + facet_wrap("~category")
)
```

:::

::: {.column width="50%"}

**`facet_grid()`** - creates a grid of panels

```{python}
#| eval: false
from plotnine import facet_grid

(
    ggplot(df, aes(x="x", y="y"))
    + geom_point()
    + facet_grid("row_var ~ col_var")
)
```

:::

:::

## Coordinate systems

Coordinate functions specify the plot's coordinate system:

```{python}
#| eval: false
from plotnine import coord_fixed, coord_cartesian, coord_trans

# Equal aspect ratio
ggplot(...) + coord_fixed()

# Zoom without clipping data
ggplot(...) + coord_cartesian(xlim=(0, 10))

# Transformed cartesian coordinate system
ggplot(...) + coord_trans()
```

## Themes {.smaller}

Themes control non-data visual elements like fonts, colors, and grid lines.

**Pre-built themes:**

```{python}
#| eval: false
from plotnine import theme_minimal, theme_bw, theme_classic

ggplot(...) + theme_minimal()
ggplot(...) + theme_bw()
ggplot(...) + theme_classic()
```

. . .

**Custom theme adjustments:**

```{python}
#| eval: false
from plotnine import theme, element_text, element_line

(
    ggplot(...)
    + theme_minimal()
    + theme(
        axis_text=element_text(size=12),
        legend_position="bottom",
        figure_size=(10, 6)
    )
)
```

## Labels

Use `labs()` to add titles and axis labels:

```{python}
#| eval: false
from plotnine import labs

(
    ggplot(df, aes(x="x_var", y="y_var"))
    + geom_point()
    + labs(
        title="My Plot Title",
        subtitle="A descriptive subtitle",
        x="X Axis Label",
        y="Y Axis Label",
        color="Legend Title"
    )
)
```

## Putting it all together {.smaller}

::: {.columns}

::: {.column width="60%"}

```{python}
#| label: fig-plotnine-example
#| fig-show: hide
from plotnine import *
from plotnine.data import mpg

(
    ggplot(mpg, aes(x="cty", y="hwy"))
    + geom_point(aes(color="displ"), alpha=0.7)
    + geom_smooth(method="lm", color="blue")
    + scale_color_continuous(cmap_name="viridis")
    + facet_wrap("~drv", ncol=1)
    + labs(
        title="City vs Highway MPG",
        x="City MPG",
        y="Highway MPG",
        color="Engine\nDisplacement"
    )
    + theme_bw()
    + theme(
        figure_size=(3, 6), 
        legend_position="bottom"
    )
)
```

:::

::: {.column .fragment width="40%"}

```{python}
#| ref-label: fig-plotnine-example
#| echo: false
```

:::

:::




## Key differences from ggplot2 {.smaller}

| ggplot2 (R) | plotnine (Python) |
|-------------|-------------------|
| `aes(x = var)` | `aes(x="var")` (quoted strings) |
| `+` at end of line | `+` at start of line (inside parens) |
| `theme(legend.position = ...)` | `theme(legend_position=...)` (underscores) |
| No parens needed | Wrap in `()` for multi-line plots |
| `ggsave()` | `.save()` method on plot object |

. . .

<br/>

```{python}
#| eval: false
# Saving a plot
p = ggplot(...) + geom_point()
p.save("my_plot.png", width=10, height=6, dpi=300)
```
