---
title: Data visualization with Python |
subtitle: Lecture 22
format: revealjs
filters: 
  - ../remove-fmt-skip.lua
---

# Warm up

## Announcements {.smaller}

- ...

## From last time: Generative art resources

-   R packages:

    -   [aRtsy](https://koenderks.github.io/aRtsy/) by Koen Derks
    -   [generativeart](https://github.com/cutterkom/generativeart) by Katharina Brunner
    -   [flametree](https://flametree.djnavarro.net/) and [jasmines](https://jasmines.djnavarro.net/) bu Danielle Navarro
    -   [ambient](https://ambient.data-imaginist.com/) and [particles](https://github.com/thomasp85/particles) by Thomas Lin Pedersen

-   More aRtists: [rtistry art gallery](https://ijeamaka-a.shinyapps.io/rtistry_gallery/) by Ijeamaka Anyene

-   A whole course on Generative Art by Danielle Navarro: [https://art-from-code.netlify.app](https://art-from-code.netlify.app/)

## Setup {.smaller}

- R:

```{r}
#| message: false
library(tidyverse)
```

- Python:

```{python}
import polars as pl
from plotnine import ggplot, aes, geom_point, geom_segment, labs, theme, theme_minimal
```

# Data visualization with Python

## Spot the difference

::: {.columns}

::: {.column}

**Plot A:**

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 4
#| echo: false
df_r <- read_csv("data/mean-area-decade.csv")

ggplot(
  df_r,
  aes(x = mean_area, y = decade_built_cat)
) +
  geom_segment(
    aes(
      x = 0,
      xend = mean_area,
      y = decade_built_cat,
      yend = decade_built_cat
    )
  ) +
  geom_point(size = 5, color = "lightblue") +
  labs(
    x = "Mean area (square feet)",
    y = "Decade built",
    title = "Mean area of houses in Duke Forest, by decade built"
  ) +
  theme_minimal()
```

:::

::: {.column}

**Plot B:**

```{python}
#| echo: false
df_py = pl.read_csv("data/mean-area-decade.csv")

(
    ggplot(df_py, aes(x="mean_area", y="decade_built_cat"))
    + geom_segment(
        aes(x=0, xend="mean_area", y="decade_built_cat", yend="decade_built_cat")
    )
    + geom_point(size=5, color="lightblue")
    + labs(
        x="Mean area (square feet)",
        y="Decade built",
        title="Mean area of houses in Duke Forest, by decade built",
    )
    + theme_minimal()
    + theme(figure_size=(8, 4))
)
```

:::

:::

## Spot the difference

::: {.columns}

::: {.column}

**Plot A:**

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 4
#| echo: false
df_r <- read_csv("data/mean-area-decade.csv")

ggplot(
  df_r,
  aes(x = mean_area, y = decade_built_cat)
) +
  geom_segment(
    aes(
      x = 0,
      xend = mean_area,
      y = decade_built_cat,
      yend = decade_built_cat
    )
  ) +
  geom_point(size = 5, color = "lightblue") +
  labs(
    x = "Mean area (square feet)",
    y = "Decade built",
    title = "Mean area of houses in Duke Forest, by decade built"
  ) +
  theme_minimal()
```

::: center-align

{{< fa brands r-project size=Huge >}}

:::

:::

::: {.column}

**Plot B:**

```{python}
#| echo: false
df_py = pl.read_csv("data/mean-area-decade.csv")

(
    ggplot(df_py, aes(x="mean_area", y="decade_built_cat"))
    + geom_segment(
        aes(x=0, xend="mean_area", y="decade_built_cat", yend="decade_built_cat")
    )
    + geom_point(size=5, color="lightblue")
    + labs(
        x="Mean area (square feet)",
        y="Decade built",
        title="Mean area of houses in Duke Forest, by decade built",
    )
    + theme_minimal()
    + theme(figure_size=(8, 4))
)
```

::: center-align

{{< fa brands python size=Huge >}}

:::

:::

:::



## Overview

::: incremental

- Goal: 
  - Get a taste of data visualization with Python with `plotnine` -- a Python data visualization package based on the grammar of graphics and inspired by `ggplot2`.
  - Not a comprehensive introduction to Python or even to data visualization with Python.

- Approach: 
  - We will cover the basics of `plotnine` and how to create different types of plots.
  - We will use `polars` to read in data from CSV files, but we will **not** cover how to prepare your data for visualization in Python.

:::

## But...

<br> <br>

::: large

[there are some Python details we can't avoid...]{.hand}

:::

# Package management <br> in Python with `uv`

## Python packaging landscape

Python's packaging ecosystem has historically been fragmented:

* Multiple tools: `pip`, `virtualenv`, `venv`, `conda`, `poetry`, `pipenv`, etc.
* Multiple config files: `requirements.txt`, `setup.py`, `pyproject.toml`, etc.
* Version management often handled separately (`pyenv`)

. . .

**uv** is a modern tool that aims to unify these concerns with a fast, Rust-based implementation.


## What is uv?

**uv** is a Python package and project manager developed by [Astral](https://astral.sh).

<br/>

Key features:

* Extremely fast (10-100x faster than pip)
* Manages Python versions
* Creates and manages virtual environments
* Installs packages
* Handles project dependencies via `pyproject.toml`
* Drop-in replacement for `pip` and `virtualenv`
* Directly supported by Positron and Reticulate

## Installing uv

**uv** is already installed on the departmental servers, for local installs:


On MacOS/Linux:

::: {.small}

```bash
curl -LsSf https://astral.sh/uv/install.sh | sh
```

:::

or with homebrew:

::: {.small}

```bash
brew install uv
```

:::

or with pip / pipx

::: {.small}

```bash
pipx install uv
pip install uv
```

:::


## Verify installation

Once installed you should be able to run the following,

```{bash}
uv --version
```

. . .

<br/>

As long as you have something higher than `0.9.*` you should be fine.

## Python versions

```{bash}
uv python list
```

## Managing Python versions

uv can install and manage multiple Python versions,

```{bash}
uv python install 3.14
```

```{bash}
uv python pin 3.14
```

. . .

The pinned version is stored in `~/.python-version` and will be used automatically.


## Initializing a project

Use `uv init` to create a new project,

::: {.small}

```bash
mkdir my-project
cd my-project
uv init
```

```
Initialized project `my-project`
```

```bash
ls -la
```

```
total 32
drwxr-xr-x   8 mine  staff   256 Apr  6 00:22 .
drwx------@ 55 mine  staff  1760 Apr  6 00:22 ..
drwxr-xr-x@  9 mine  staff   288 Apr  6 00:22 .git
-rw-r--r--@  1 mine  staff   109 Apr  6 00:22 .gitignore
-rw-r--r--@  1 mine  staff     5 Apr  6 00:22 .python-version
-rw-r--r--@  1 mine  staff    88 Apr  6 00:22 main.py
-rw-r--r--@  1 mine  staff   156 Apr  6 00:22 pyproject.toml
-rw-r--r--@  1 mine  staff     0 Apr  6 00:22 README.md
```

:::

. . .

This creates a `pyproject.toml`, a sample `main.py` script, and basic git infrastructure. Generally, we only really care about the `pyproject.toml` which we can exclusively generate via `uv init --bare`.

## `pyproject.toml`

Modern project metadata file, tracks python version and package dependencies among other details.

```toml
[project]
name = "my-project"
version = "0.1.0"
description = "Add your description here"
readme = "README.md"
requires-python = ">=3.14"
dependencies = []
```

## Adding dependencies

Once we have our project setup we can add (and install) dependencies directly via uv. `uv add` updates `pyproject.toml` and installs the package (creating a venv if needed).

::: {.columns .xsmall}

::: {.column}

```bash
uv add plotnine
```

```
Using CPython 3.14.3
Creating virtual environment at: .venv
Resolved 19 packages in 188ms
Installed 17 packages in 134ms
 + contourpy==1.3.3
 + cycler==0.12.1
 + fonttools==4.62.1
 + kiwisolver==1.5.0
 + matplotlib==3.10.8
 + mizani==0.14.4
 + numpy==2.4.4
 + packaging==26.0
 + pandas==3.0.2
 + patsy==1.0.2
 + pillow==12.2.0
 + plotnine==0.15.3
 + pyparsing==3.3.2
 + python-dateutil==2.9.0.post0
 + scipy==1.17.1
 + six==1.17.0
 + statsmodels==0.14.6
```

:::

::: {.column .fragment}

```bash
uv add polars
```

```
Resolved 21 packages in 142ms
Installed 2 packages in 6ms
 + polars==1.39.3
 + polars-runtime-32==1.39.3
```

:::

::: 

## Updated `pyproject.toml`

```toml
[project]
name = "my-project"
version = "0.1.0"
description = "Add your description here"
readme = "README.md"
requires-python = ">=3.14"
dependencies = [
    "plotnine>=0.15.3",
    "polars>=1.39.3",
]
```

## Virtual environments

Virtual environments isolate project dependencies from the system Python and other projects. Packages are installed in a local folder in your project.

As we just saw, using `uv add` will create a new virtual environment in `.venv` by default if there is not an existing venv.

## Activating environments in Positron with uv

Positron automatically detects virtual environments in your project directory. When you open a folder containing a `.venv` directory (created by uv), Positron will:

* Detect the environment and offer to use it
* Show the active Python interpreter in the status bar
* Use the environment for the Python console and when running scripts

If not automatically detected, you can manually select the interpreter via the Command Palette (`Cmd+Shift+P` / `Ctrl+Shift+P`) and searching for "Python: Select Interpreter".

## `uv sync`

Since the `.venv` folder is system specific (and large) it is not typically committed to git. Instead you will likely clone a repository that just has a `pyproject.toml` file.

Use `uv sync` to construct the venv and install all dependencies for the project 

## Listing packages

::: {.small}

```bash
uv pip list
```

```
Package           Version
----------------- -----------
contourpy         1.3.3
cycler            0.12.1
fonttools         4.62.1
kiwisolver        1.5.0
matplotlib        3.10.8
mizani            0.14.4
numpy             2.4.4
packaging         26.0
pandas            3.0.2
patsy             1.0.2
pillow            12.2.0
plotnine          0.15.3
polars            1.39.3
polars-runtime-32 1.39.3
pyparsing         3.3.2
python-dateutil   2.9.0.post0
scipy             1.17.1
six               1.17.0
statsmodels       0.14.6
```

```bash
uv pip show plotnine
```

```
Name: plotnine
Version: 0.15.3
Location: /Users/mine/Desktop/my-project/.venv/lib/python3.14/site-packages
Requires: matplotlib, mizani, numpy, pandas, scipy, statsmodels
Required-by:
```

:::

`uv pip install` and `uv pip freeze` are also available as lower-level commands for working outside of a project context or with `requirements.txt` files.


## Dependency tree

::: {.xsmall}

```bash
uv tree
```

```
Resolved 21 packages in 18ms
my-project v0.1.0
├── plotnine v0.15.3
│   ├── matplotlib v3.10.8
│   │   ├── contourpy v1.3.3
│   │   │   └── numpy v2.4.4
│   │   ├── cycler v0.12.1
│   │   ├── fonttools v4.62.1
│   │   ├── kiwisolver v1.5.0
│   │   ├── numpy v2.4.4
│   │   ├── packaging v26.0
│   │   ├── pillow v12.2.0
│   │   ├── pyparsing v3.3.2
│   │   └── python-dateutil v2.9.0.post0
│   │       └── six v1.17.0
│   ├── mizani v0.14.4
│   │   ├── numpy v2.4.4
│   │   ├── pandas v3.0.2
│   │   │   ├── numpy v2.4.4
│   │   │   └── python-dateutil v2.9.0.post0 (*)
│   │   └── scipy v1.17.1
│   │       └── numpy v2.4.4
│   ├── numpy v2.4.4
│   ├── pandas v3.0.2 (*)
│   ├── scipy v1.17.1 (*)
│   └── statsmodels v0.14.6
│       ├── numpy v2.4.4
│       ├── packaging v26.0
│       ├── pandas v3.0.2 (*)
│       ├── patsy v1.0.2
│       │   └── numpy v2.4.4
│       └── scipy v1.17.1 (*)
└── polars v1.39.3
    └── polars-runtime-32 v1.39.3
(*) Package tree already displayed
```

:::


## Common workflows

New project setup:

::: {.small}

```bash
mkdir my-project # or use Positron's new project wizard
cd my-project    # and navigate to the project folder
uv init --bare
uv add plotine polars
```

:::

. . .

<br/>

Clone existing project:

::: {.small}

```bash
git clone <repo-url> # or use Positron's git integration
cd <repo>            # and navigate to the project folder
uv sync
```

:::

## Let's give it a try!

::: task
Go to `ae-16` and let's make a simple plot with `plotnine`!
:::
